---
title: "weightloss_analysis"
author: "Mikkel Skjoldan Svenningsen"
date: "2026-03-16"
output: html_document
---

# Analysis of repeated-measures weightloss proteomics  
This dataset contains plasma samples from 102 patients prior to and following weightloss. It is a repeated-measures longitudinal dataset: each patient were sampled at -8,0,4,13,26,39, and 52 weeks. Since there are 7 measurements from each patient, the measurements are not independent and should be analyzed with repeated-measures statistics. First the publicly data is imported and prepared for analysis. 

```{r import data and metadata, include=FALSE}

library(nlme)
library(lme4)
library(lmerTest)
library(tidyr)
library(readxl)
library(LMMstar)
library(ggplot2)
library(dplyr)
library(Hmisc)
library(here)

here::i_am("in_silico_analysis/6_supplement_missing_values.Rmd")

M <- read_excel("D:/Data_PRIDE/Mann_obesity (PXD004242)/data_modified.xlsx", sheet="Meta")
D <- read_excel("D:/Data_PRIDE/Mann_obesity (PXD004242)/data_modified.xlsx", sheet="Data_sub")

t_in <- c(1,2,3,4,5,6,7)
w_in <- c(-8, 0, 4, 13, 26, 39, 52)

uid2gene <- setNames(D$`Majority protein IDs`, D$`Gene names`)

time2week <- w_in
names(time2week) <- t_in

person <- c()
time <- c()
bmi <- c()
height <- c()
weight <- c()
for(col in colnames(M[, -1])){
  a <- strsplit(col, "_")[[1]]
  person <- c(person, a[2])
  time <- c(time, unname(time2week[a[3]]))
  weight <- c(weight, as.numeric(M[1, col]))
  bmi <- c(bmi, as.numeric(M[2, col]))
  height <- c(height, sqrt(as.numeric(M[1, col])/as.numeric(M[2, col])))
}

df_meta <- data.frame( person, time, weight, bmi, height)

df_meta <- df_meta %>%
  group_by(person) %>%
  mutate(height_estimate = mean(height, na.rm = TRUE)) %>%
  as.data.frame()

df_meta$bmi_calc <- (df_meta$weight)/(df_meta$height_estimate^2)


df_meta$bmi_cat <- cut2(df_meta$bmi_calc, g = 4)

df_in <- data.frame(D)

df_dat <- df_in[, c(-1,-2,-3)]

rownames(df_dat) <- df_in$Majority.protein.IDs

df_dat[df_dat=="NaN"] <- NA 
```

The repeated-measures statistical analysis. 


```{r setup, include=FALSE}

df_fit <- data.frame(dv = dim(df_meta)[1], 
                     time = factor(as.numeric(df_meta$time)), 
                     person = df_meta$person, 
                     weight = df_meta$weight, 
                     weight_binary = ifelse(df_meta$weight<87.05, 1, 0), 
                     bmi = df_meta$bmi_calc, 
                     bmi_cat = factor(as.integer(df_meta$bmi_cat)))

rows <- rownames(df_dat)

safe_extract <- function(expr){
  tryCatch(expr, error = function(e) NA)
  
}

time_labels <- c("time0", "time4", "time13", "time26", "time39", "time52")

M_time <- combn(time_labels, 2)

time_list1 <- M_time[1, ]
time_list2 <- M_time[2, ]


res <- lapply(rows, function(row){
  
  y <- as.numeric(df_dat[row, ])
  print(row)
  if(mean(!is.na(y))<0.7){ return(NULL) }
    
  ind_real <- !is.na(y)
  df_fit$dv <- y
  
  
  fit <- try(lmm(dv ~ time, repetition = ~time|person, structure = "UN", data = df_fit[ind_real, ]))
  
  print(which(rows == row))
  
  q <- try(anova(fit))
  
  b_success1 <- !inherits(fit, "try-error")
  b_success2 <- !inherits(q, "try-error")
  
  if(!b_success1 | !b_success2) return(NULL)

  
  q <- anova(fit)
  lop <- list()
  lopar <- list()
  
  df_coef <- data.frame(coef(fit))
  
  for(time_label_i in time_labels){
    lop[[time_label_i]] <- safe_extract( anova(fit, effects = c(paste0(time_label_i, "=0")))$multivariate$p.value)
  }
  
  
  for(time_label_i in time_list1){
    
    for(time_label_j in time_list2){
   
      if(time_label_i != time_label_j){
        lop[[paste0(time_label_i, "vs", time_label_j)]] <- safe_extract( anova(fit, effects = c(paste0(time_label_i,"-", time_label_j, "=0")))$multivariate$p.value)
      }
    }
  }
  
  for(time_label_i in time_labels){
    lopar[[paste0("par_", time_label_i)]] <- df_coef[rownames(df_coef)==time_label_i, ]
  }
  
   for(time_label_i in time_list1){
    
    
    for(time_label_j in time_list2){
   
      if(time_label_i != time_label_j){
        lopar[[paste0("par_", time_label_i, "vs", time_label_j)]] <- df_coef[rownames(df_coef)==time_label_j, ] - df_coef[rownames(df_coef)==time_label_i, ]
      }
    }
  }
    
    
      
    df_out <- data.frame(lop)
    
    gn <-  D[D$`Majority protein IDs`==row, ]$`Gene names`
    
    print(gn)
    
    df_out$gene <- gn
    df_out$uid <- row
    
    df_out <- cbind(df_out, data.frame(lopar))

  return(df_out)
    
    
}) # lapply 


df_results <- do.call(rbind, res)


list_fdrs <- list()
  
for(time_label_i in time_list1){

    list_fdrs[[paste0("fdr_",time_label_i)]] <- p.adjust(df_results[[time_label_i]], method = "BH")
  
  
  for(time_label_j in time_list2){
 
    if(time_label_i != time_label_j){
      list_fdrs[[paste0("fdr_",time_label_i,"vs", time_label_j)]] <- p.adjust(df_results[[paste0( time_label_i,"vs", time_label_j)]], method = "BH")
    }
  }
}


df_plot <- cbind(df_results, as.data.frame(list_fdrs))

df_sub <- df_plot %>%
  dplyr::select(contains("fdr_t")) %>% 
  mutate(min_fdr_t = apply(., 1, min, na.rm = TRUE))

df_plot$min_fdr_t <- df_sub$min_fdr_t
df_plot$sig_tt <- df_plot$min_fdr_t<0.05

saveRDS(df_plot, file = "D:/Project_longitudinal_methods/Manuscript/Github_structure/Results_real_world/weightloss/df_regression_onlyTime_.rds")

```



```{r find trajectories }



data_path <- here(  "Results_real_world", "weightloss")

df_plot <- readRDS(file = paste0(data_path, "/df_regression_onlyTime.rds"))

### add new significance criterium

df_sub <- df_plot %>%
  dplyr::select(contains("fdr_t")) %>% 
  dplyr::select(-contains("vs")) %>%
    dplyr::select(-contains("min")) %>%
  mutate(min_fdr_t_vs0 = apply(., 1, min, na.rm = TRUE))

df_plot$min_fdr_t_vs0 <- df_sub$min_fdr_t_vs0
df_plot$sig_tt_vs0 <- df_plot$min_fdr_t_vs0<0.05

ind_sig <- df_plot$sig_tt_vs0
ind <- rownames(df_dat) %in% df_plot[ind_sig, ]$uid

df_par <- df_plot[ind_sig, ] %>%
  select(par_time0, par_time4, par_time13, par_time26, par_time39, par_time52)

rownames(df_par) <-df_plot[ind_sig, ]$uid

df_sc <- t(scale(t(df_par)))

data_path <- here(  "Figures")

p <- pheatmap(df_sc, cutree_rows = 6, 
              filename = paste0(data_path, "/pheatmap_weightloss.pdf"), 
              show_rownames = FALSE, show_colnames = FALSE)

ord <- p$tree_row$order
clusters <- cutree(p$tree_row, k=6)
arrangement <- unique(clusters[ord])


term_clusters <- data.frame(
  Cluster = clusters, 
  order = ord
)


count <- 0
dd <- list()
lop <- list()
for(cluster_i in unique(clusters)){
  
  count <- count + 1
  
  y <- c()
  ti <- c()
  patient <- c()
  names <- rownames(term_clusters[term_clusters$Cluster==cluster_i, ])
  nn <- c()
  for(name_i in names){
    y <- c(y, c(0, as.numeric(df_sc[name_i, ])))
    ti <- c(ti, c(-8,0,4,13,26,39,52))
    nn <- c(nn, rep(name_i,7))

  }
  
  df_plot <- data.frame(y,ti, nn)

  df_plot$nn <- factor(df_plot$nn, levels = unique(df_plot$nn))
  df_plot$ti <- factor(df_plot$ti, levels = c(-8,0,4,13,26,39,52))
  
   dd[[cluster_i]] <- df_plot

   
  df_me <- df_plot %>%
  group_by(ti) %>%
  dplyr::summarise(mean_y = mean(y, na.rm = TRUE)) %>%
  mutate(ti = factor(ti, levels = levels(df_plot$ti)))
   
if(count == 0){
  g <- ggplot(df_plot) + 
    geom_line(aes(ti, y, group = nn), alpha = 0.1) + 
    geom_line(data = df_me, aes(ti, mean_y, group = 1), color = "red", size = 1.2) +  # Add mean trajectory
    labs(x = "", y = "standardized mean protein intensity") + 
    #scale_x_discrete(labels = c("A" = "Tumor-pre", "B" = "Tumor-post", "C" = "Normal")) + 
    theme_minimal() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12 ), 
          panel.grid.major.y = element_blank(),
          panel.grid.minor = element_blank())
  
} else {
  g <- ggplot(df_plot) + 
    geom_line(aes(ti, y, group = nn), alpha = 0.1) + 
    geom_line(data = df_me, aes(ti, mean_y, group = 1), color = "red", size = 1.2) +  # Add mean trajectory
    labs(x = "", y = NULL) + 
    #scale_x_discrete(labels = c("A" = "Tumor-pre", "B" = "Tumor-post", "C" = "Normal")) + 
    theme_minimal() + 
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.title.y = element_blank(), 
          axis.text.x = element_text(angle = 45, hjust = 1, size = 12), 
         panel.grid.major.y = element_blank(),
          panel.grid.minor = element_blank())
}
  
  
  lop[[cluster_i]] <- g
 
}


g <- ggarrange(lop[[1]], lop[[2]], lop[[3]], lop[[4]], lop[[5]],lop[[6]], ncol = 6)


data_path <- here(  "Figures")

ggsave(plot = g, paste0(data_path, "/obesity_trajectories_.pdf"), width=8, height=3)


```

```{r example functional analysis of trajectory clusters}

source("D:/Automated_functional_analysis/R/functions_functional_analysis.R")

### tmp

universe <- D$`Gene names`

uid2gene <- setNames(D$`Gene names`, D$`Majority protein IDs`)

list_of_ora_art <- list()
for(i in unique(clusters)){

  
  ind <- clusters == i 
  names_sub <- names(clusters)[ind]
  
  genes_sub <- c()
  for(name in names_sub){
    genes_sub <- c(genes_sub, uid2gene[[name]])
  }
  
  folder_databases <- "N:/SUN-BMI-CardiacProteomics/Scripts/Mikkel/automated-functional-analysis"
  
  database <- get_database( "reactome_human", folder_databases )
  
  out1 <- genes_ORA(genes_sub, universe, database, 5,500)

  list_of_ora_art[[i]] <- out1[out1$pv_bh<=0.2, ]$functional_category
  
  ora_plot <- out1[out1$pv_bh<=0.2, ]

  ora_plot <- ora_plot[order(ora_plot$gene_ratio), ]
  
  
  print(ora_plot)
  #ora_plot$pw_nice <- substr(ora_plot$pw_nice, 1, 50)
  #ora_plot$pw_nice <- factor(ora_plot$pw_nice, levels = ora_plot$pw_nice)
  
  #g <- ggplot(ora_plot, aes(x = gene_ratio, y = pw, size = gene_count, color = pv_bh)) +
  #  geom_point() +
  #  scale_color_gradient(low = "red", high = "blue") +
  #  theme_bw() +
  #  labs(title = paste0("ORA ART cluster:", i, "_size:", sum(ind)), x = "Gene Ratio", y = "Description", color = "Adjusted P-value", size = "Count")
  
  #ggsave(filename = paste0("E:/Project_longitudinal_methods/Figures/Geiger ORA/cluster_art_",i, ".png"), plot = g)
  
}





```


